---
title: "Shift Share"
author: "Joe Yuke"
date: "12/31/2020"
output: html_document
---

<!-- comment -->

|     Shift-Share analysis is a technique used by urban planners, regional analysts, and economic developers to identify the sources of regional economic growth. Assuming job growth in Santa Barbara County can be explained by three components: National Share, Industry Mix, and Regional Shift, we can compare different sectors of the economy and identify competitive industries as the ones that outperform their counterpart at the state/national level. This information can then be used to inform public investment decisions and economic development policy.


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width= 8, fig.path='Output/', echo = TRUE)
```

```{r, include=FALSE}
# 6 input vars:
#                                     "b" for base year
# local jobs in industry i @ t-1:     local_i_b
# local jubs in industry i @ t:       local_i
# national jobs @ t-1:                US_b
# national jobs @ t:                  US
# national jobs in industry i @ t-1:  US_i_b
# national jobs in industry i @ t:    US_i


# SS = NS + IM + RS
# national share:   local_i_b*(US/US_b)
# Industry mix:     (local_i_b*US_i/US_i_b) - NS
# Regional shift:   local_i_b*(local_i/local_i_b - US_i/US_i_b)


# ------------------------- #
#                           #
#       Preliminaries       #
#                           #
# ------------------------- #

# packages
library(xlsx)
library(tidyverse)
library(ggplot2)
library(scales)

# directories, manually set
user           <- "C:/Users/Joe Yuke/OneDrive - BAY AREA TECHNOLOGY MANAGEMENT INC/Documents/EFP/Shift_share_proj/"
county_folder  <- "SBC"

pwd            <- paste0(user,county_folder)
setwd(pwd)
outputloc   <- paste0(pwd,"/Output")

# color template
EFPcolors   <- c("deepskyblue4","brown3","darkolivegreen4", "cornflowerblue",
                 "darkorange3","brown4","mediumpurple4","red", "darkslateblue")



# ---------------- #
#                  #
#       Main       #
#                  #
# ---------------- #

# manually set parameters
base_yr <- 2018
yr      <- 2019
County  <- "Santa Barbara"


# industries of interest: NAICS code
code_list <- c("11", "23", "31-33", "42", "44-45", "48-49", "51", "52", "53",
               "54", "55", "56", "62", "71", "72")
ind_names <- c("Agriculture", "Construction", "Manufacturing", 
               "Wholesale Trade", "Retail Trade", 
               "Transportation & Warehousing", "Information",
               "Finance & Insurance", "Real Estate", 
               "Prof. & Technical Services", "Management", 
               "Admin & Waste Services", "Health Care & \nSocial Assistance",
               "Arts, Entertainment, & Recreation", 
               "Accommodation & \nFood Services")
num_i     <- length(code_list)


# shift share function:
shiftshare <- function(li1,li2,ri1,ri2,r1,r2){
  
  # "l" for local; "r" for regional
  # 1 for time period 1; 2 for time period 2
  NS <- round(li1 * (r2/r1))
  IM <- round((li1 * ri2/ri1) - NS)
  RS <- round(li1 * (li2/li1 - ri2/ri1))
  return(c(NS,IM,RS))
  
}


# Data from: State and County Employment and Wages
# (Quarterly Census of Employment & Wages - QCEW)
# https://www.bls.gov/data/
# Note: Private Sector Jobs

# Totals for national/state/local:
US_data    <- read.xlsx("National/US_tot.xlsx", sheetIndex=1, startRow=13)
US_data    <- select(US_data, Year, Annual) %>% subset(Year == base_yr | Year == yr)
CA_data    <- read.xlsx("State/CA_tot.xlsx", sheetIndex=1, startRow=13)
CA_data    <- select(CA_data, Year, Annual) %>% subset(Year == base_yr | Year == yr)
local_data <- read.xlsx("Local/loc_tot.xlsx", sheetIndex=1, startRow=13)
local_data <- select(local_data, Year, Annual) %>% subset(Year == base_yr | Year == yr)

# Picking out data points
US_b    <- US_data[1,2]
US      <- US_data[2,2]
CA_b    <- CA_data[1,2]
CA      <- CA_data[2,2]
local_b <- local_data[1,2]
local   <- local_data[2,2]

# county/US matrix
US_Shift_share <- matrix(data = 0,num_i,4)
colnames(US_Shift_share)[1:4] <- c("Industry(NAICS)","National_Share",
                                   "Industry_Mix","Regional_Share")
US_Shift_share[,1] <- code_list
rownames(US_Shift_share)[1:num_i] <- ind_names


# county/CA matrix
CA_Shift_share <- matrix(data = 0,num_i,4)
colnames(CA_Shift_share)[1:4] <- c("Industry(NAICS)","CA_Share",
                                   "Industry_Mix","Regional_Share")
CA_Shift_share[,1] <- code_list
rownames(CA_Shift_share)[1:num_i] <- ind_names


# industry employment changes matrix
e_change <- matrix(data = 0,num_i,10)
colnames(e_change)[1:10] <- c("Industry(NAICS)","t-1:county","t:county",
                              "Pct_Chng:county","t-1:state","t:state","Pct_Chng:state",
                              "t-1:national","t:national","Pct_Chng:national")
e_change[,1] <- code_list
rownames(e_change)[1:num_i] <- ind_names

# shift-share calculations:
for (i in 1:num_i) {
  
  US_i_data <- read.xlsx(sprintf("National/US_%s.xlsx",code_list[i]), sheetIndex = 1, startRow = 13)
  US_i_data <- select(US_i_data, Year, Annual) %>%
    subset(Year == base_yr | Year == yr)
  CA_i_data <- read.xlsx(sprintf("State/CA_%s.xlsx",code_list[i]), sheetIndex = 1, startRow = 13)
  CA_i_data <- select(CA_i_data, Year, Annual) %>%
    subset(Year == base_yr | Year == yr)
  loc_i_data <- read.xlsx(sprintf("Local/loc_%s.xlsx",code_list[i]), sheetIndex = 1, startRow = 13)
  if ("" %in% loc_i_data$Annual) {
    loc_i_data$Annual <- as.integer(as.character(loc_i_data$Annual))
  }
  loc_i_data <- select(loc_i_data, Year, Annual) %>%
    subset(Year == base_yr | Year == yr)
  
  #Picking out data points:
  US_i_b    <- US_i_data[1,2]
  US_i      <- US_i_data[2,2]
  CA_i_b    <- CA_i_data[1,2]
  CA_i      <- CA_i_data[2,2]
  local_i_b <- loc_i_data[1,2]
  local_i   <- loc_i_data[2,2]
  
  #using shift-share function:
  US_Shift_share[i,2:4] <- shiftshare(local_i_b,local_i,US_i_b,US_i,US_b,US)
  CA_Shift_share[i,2:4] <- shiftshare(local_i_b,local_i,CA_i_b,CA_i,CA_b,CA)
  
  #percent change:
  US_chng    <- round((US_i - US_i_b)/US_i_b, digits = 3)*100
  CA_chng    <- round((CA_i - CA_i_b)/CA_i_b, digits = 3)*100
  local_chng <- round((local_i - local_i_b)/local_i_b, digits = 3)*100
  e_change[i,2:10] <- c(local_i_b,local_i,local_chng,CA_i_b,CA_i,CA_chng,
                        US_i_b,US_i,US_chng)
  
}

# Total, all industries pct change
US_chng    <- round((US - US_b)/US_b, digits = 3)*100
CA_chng    <- round((CA - CA_b)/CA_b, digits = 3)*100
local_chng <- round((local - local_b)/local_b, digits = 3)*100

`Total Private Employment` <- c("",local_b,local,local_chng,
                                CA_b,CA,CA_chng,
                                US_b,US,US_chng)

# NOT FOR GRAPHING
e_change1 <- rbind(e_change,`Total Private Employment`)


# Save Output as an excel file:
write.xlsx(US_Shift_share, file=paste(outputloc,"Shift_share.xlsx",sep="/"), sheetName="US", append=F)
write.xlsx(CA_Shift_share, file=paste(outputloc,"Shift_share.xlsx",sep="/"), sheetName="CA", append=T)
write.xlsx(e_change1, file=paste(outputloc,"Shift_share.xlsx",sep="/"), sheetName="Change", append=T)


#--------------#
#              #
#     Plots    #
#              #
#--------------#

Years       <- c((yr-9):yr)

# number of industries to graph
num_j       <- 3


# data frame for plots
graph_data <- as.data.frame(matrix(data = 0, length(Years), 4+3*num_j))
colnames(graph_data)[1] <- "Year"
colnames(graph_data)[(2+3*num_j):(4+3*num_j)] <- c("US_tot","State_tot","Local_tot")


# Identifying n largest local industries:

# manual input option:
#graph_codes <- c("11", "62", "72")
#graph_names <- c("Agriculture","Health Care & Social Assistance", "Accommodation & Food Services")

graph_codes <- rep(0,num_j)
graph_names <- rep(0,num_j)

ind_size <- as.data.frame(e_change) %>% select(`Industry(NAICS)`,`t:county`)
ind_size$`t:county` <- as.integer(as.character(ind_size$`t:county`))
ind_size[is.na(ind_size)] <- 0
ind_size1 <- ind_size

for (j in 1:num_j) {
  
  index <- which.max(ind_size1$`t:county`)
  graph_codes[j] <- ind_size1$`Industry(NAICS)`[index]
  
  ind_size1 <- ind_size1[-index,]
  
}

graph_codes <- sort(as.integer(graph_codes))

for (j in 1:num_j) {
  
  index <- as.integer(graph_codes[j])
  
  graph_codes[j] <- as.character(ind_size$`Industry(NAICS)`[as.integer(graph_codes[j])])
  
  graph_names[j] <- row.names(ind_size)[index]
  
  colnames(graph_data)[(3*j-1):(3*j+1)] <- 
    sprintf(c("US_%s","State_%s","Local_%s"), graph_codes[j])
  
}


# Filling in graph_data:
graph_data$Year <- Years
graph_data$Year <- as.Date(paste0(as.factor(graph_data$Year),"-01-01"))

for (j in 1:num_j) {
  
  US <- read.xlsx(sprintf("National/US_%s.xlsx",graph_codes[j]), sheetIndex = 1, startRow = 13) %>%
    select(Year, Annual)
  graph_data[,(3*j-1)] <- US$Annual[1:10]
  
  State <- read.xlsx(sprintf("State/CA_%s.xlsx",graph_codes[j]), sheetIndex = 1, startRow = 13) %>%
    select(Year, Annual)
  graph_data[,(3*j)] <- State$Annual[1:10]
  
  Local <- read.xlsx(sprintf("Local/loc_%s.xlsx",graph_codes[j]), sheetIndex = 1, startRow = 13) %>%
    select(Year, Annual)
  graph_data[,(3*j+1)] <- Local$Annual[2:11] #manual, needs adjusting based on data availibility
  
}

US <- as.data.frame(read.xlsx("National/US_tot.xlsx", sheetIndex = 1, startRow = 13))
graph_data$US_tot <- US$Annual[1:10]
State <- as.data.frame(read.xlsx("State/CA_tot.xlsx", sheetIndex = 1, startRow = 13))
graph_data$State_tot <- State$Annual[1:10]
Local <- as.data.frame(read.xlsx("Local/loc_tot.xlsx", sheetIndex = 1, startRow = 13))
graph_data$Local_tot <- Local$Annual[1:10]


# bar chart manipulation:
graph_data_bar <- gather(tail(graph_data,5), type, employment, -Year)
graph_data_bar <- separate(graph_data_bar, type, c("location","industry"), sep = "_")


# defining theme:
efp_theme <- theme( text = element_text(family = "serif"),
                    legend.title = element_blank(),
                    legend.background = element_blank(),
                    legend.key = element_blank(),
                    panel.grid.major = element_blank() ,
                    panel.grid.minor = element_blank() ,
                    plot.title = element_text(face = "bold", size = 16),
                    axis.title.y = element_text(size = 14),
                    axis.title.x = element_blank(),
                    axis.line = element_line(color = 'black'),
                    plot.subtitle = element_text(face = "italic"),
                    plot.caption = element_text(hjust = -0.3, face= "italic", size = 8),
                    legend.position = 'bottom',
                    panel.background = element_rect(fill = 'white'))

```
|
|     In 2019, Santa Barbara County added 4,249 jobs, achieving a 2.5% increase across all industries. This is a significant change from the prior year’s growth of 2,715 jobs. Notably, employment success was driven by the Professional, Scientific & Technical Services sector which grew by 633 jobs; the Health Care & Social Assistance sector’s gain of 647 jobs; and Agriculture’s increase of 1,227 jobs.
|
|     At the state and national levels, there were also increases in the total number of jobs but at less pronounced rates. California added 251,568 jobs in 2019, representing a 1.7% increase from the previous year, and the US added 1,806,905 jobs, growing by 1.5%.
|
<!-- [employment change] -->
## Employment by Industry Sector
```{r, include=FALSE}
library(knitr)
library(kableExtra)

`-`            <- c("", rep(c("2018","2019","%Change"),3))
e_change1       <- rbind(`-`, e_change1)
e_change_table <- data.frame(e_change1) %>% 
  select(-`Industry.NAICS.`)
```
```{r, echo=FALSE}
kable(e_change_table, col.names = c("","SB","",
                                    "","CA","",
                                    "","US",""),
      escape = F, caption = "2018 - 2019") %>%
  kable_styling(latex_options = "hold_position")
```
|
|     Agriculture, Health Care, and the Accommodation & Food sector have consistently been Santa Barbara County’s largest industries in terms of employment. Over the past 10 years, all three have experienced fairly steady growth as well. However, to assess industry performance, we should look beyond the raw employment numbers.
|
|
<!-- [County linegraph] -->
```{r Local_graph_ind, echo=FALSE}
ggplot(graph_data) +
  geom_line(size = 1.2, aes(x = Year, y = graph_data[,4], color = "first")) +
  geom_line(size = 1.2, aes(x = Year, y = graph_data[,7], color = "second")) +
  geom_line(size = 1.2, aes(x = Year, y = graph_data[,10], color = "third")) +
  scale_x_date(labels = date_format("%Y"),
               breaks = seq.Date(from = min(graph_data$Year), 
                                 to = max(graph_data$Year), by = "2 years")) +
  scale_y_continuous(labels = scales::comma_format()) +
  ggtitle(sprintf("%s County Employment",County),"Private Sector") +
  labs(caption = "Source: Quarterly Census of Employment and Wages (QCEW)") +
  ylab("Annual Employment") +
  scale_color_manual(labels = c('first'  =  graph_names[1],
                                'second' =  graph_names[2],
                                'third'  =  graph_names[3]),
                     values = c('first'  = EFPcolors[3],
                                'second' = EFPcolors[1],
                                'third'  = EFPcolors[2])) +
  efp_theme +
  theme(panel.grid.major.y = element_line(color="gray"))
```

|     Shift-Share analysis can explain how much of the job growth within an industry should be attributed to local conditions. As an example, Retail Trade, the county’s fourth largest industry, added a mere 2 jobs in 2019. In fact, this small increase is a competitive performance compared to the national and state retail industries. The “National Share” calculation yields 18,816 which represents the number of jobs that would exist in the local retail industry had it grown at the same rate as the whole US economy. In other words, how much job growth should have been expected according to the national business cycle. “Industry Mix” assumes changes in employment consistent with the growth of the national retail industry. The US-SB analysis identifies a negative Industry Mix of -490. This is because the Retail Trade sector performed poorly at the national level, declining by 1.2% in 2019. Finally, the “Regional Shift” identifies leading industries by finding the differential between the local and national rates of growth and using it as a multiplier for the previous year’s job numbers. The Retail Trade sector for Santa Barbara County achieved a regional shift of 223 jobs, which represents 1.2% of local retail jobs. Industries that have positive Regional Shift scores are deemed competitive according to Shift-Share analysis. Adding all three calculations together yields the true employment numbers for the industry of interest in 2019. For this report, we also run the analysis substituting California’s employment numbers for the US’s statistics. Interestingly, Santa Barbara’s Retail industry performed even stronger relative to the state of California.
|
|     The Professional, Scientific, and Technical Services industry, which encompasses lawyers, engineers, researchers etc., achieved a relatively high Regional Shift of 331, or 2.8% of employment in that industry. This sector has been increasing slightly since 2016 but has largely remained consistent for the past 10 years.

<!-- [Shift share table] -->
## Shift Share by Industry Sector
```{r, include=FALSE}
`-` <- c("","National Share", "Industry Mix", "Regional Share")
US_Shift_share <- rbind(`-`,US_Shift_share)
US_ss_table <- data.frame(US_Shift_share) %>% 
  select(-`Industry.NAICS.`)

`-` <- c("","State Share", "Industry Mix", "Regional Share")
CA_Shift_share <- rbind(`-`,CA_Shift_share)
CA_ss_table <- data.frame(CA_Shift_share) %>% 
  select(-`Industry.NAICS.`)

ss_table <- cbind(US_ss_table,CA_ss_table)
```
```{r, echo=FALSE}
kable(ss_table, col.names = c("","US-SB","","","CA-SB",""),
      escape = F, caption = "Santa Barbara: 2018 - 2019") %>%
  kable_styling(latex_options = "hold_position")
```
|	
|     Out of the County’s top three industries, Agriculture was the only one to realize a positive Regional Shift across both US and CA calculations. In the US-SB calculation, the Regional Shift was 1,236, or 5.1% of agricultural employment, making it the most competitive industry for Santa Barbara in 2019. Statewide and nationally, Agriculture represents a much smaller portion of the labor force and is typically outpaced in terms of growth when compared to Santa Barbara county. Health Care saw a modest increase of 647 jobs, but relative to the changes in California's Health Care employment, this does not represent significant growth. Finally, the Accommodation and Food Services industry added only 222 jobs and underperformed compared to the national and particularly the state level. This is indicated in the much more negative Regional Shift for the CA-SB analysis.
|
|
<!-- [Local_bar] [State_bar] [US_bar] -->
```{r Local_bar, echo=FALSE}
ggplot() +
  geom_col(data = subset(graph_data_bar, location=="Local" & industry=="tot"),
           aes(x = Year, y = employment, fill = "Total Employment"),
           width=300) +
  geom_col(data = subset(graph_data_bar, location=="Local" & industry!="tot"),
           aes(x = Year, y = employment, fill = industry), width=300) +
  scale_y_continuous(labels = scales::comma_format()) +
  ggtitle(sprintf("%s County Private Employment",County),
          "Industry Portion of Total Employment") +
  labs(caption = "Source: Quarterly Census of Employment and Wages (QCEW)") +
  ylab("Annual Employment") +
  scale_fill_manual(labels = c(graph_names[1], graph_names[2], 
                               graph_names[3], 'Total \nEmployment'),
                    values = c(EFPcolors[3], EFPcolors[1], 
                               EFPcolors[2], EFPcolors[4])) +
  efp_theme
```

```{r State_bar, echo=FALSE}
ggplot() +
  geom_col(data = subset(graph_data_bar, location=="State" & industry=="tot"),
           aes(x = Year, y = employment, fill = "Total Employment"),
           width=300) +
  geom_col(data = subset(graph_data_bar, location=="State" & industry!="tot"),
           aes(x = Year, y = employment, fill = industry), width=300) +
  scale_y_continuous(labels = scales::comma_format()) +
  ggtitle("California Private Employment",
          "Industry Portion of Total Employment") +
  labs(caption = "Source: Quarterly Census of Employment and Wages (QCEW)") +
  ylab("Annual Employment") +
  scale_fill_manual(labels = c(graph_names[1], graph_names[2], 
                               graph_names[3], 'Total \nEmployment'),
                    values = c(EFPcolors[3], EFPcolors[1], 
                               EFPcolors[2], EFPcolors[4])) +
  efp_theme
```

```{r US_bar, echo=FALSE}
ggplot() +
  geom_col(data = subset(graph_data_bar, location=="US" & industry=="tot"),
           aes(x = Year, y = employment, fill = "Total Employment"),
           width=300) +
  geom_col(data = subset(graph_data_bar, location=="US" & industry!="tot"),
           aes(x = Year, y = employment, fill = industry), width=300) +
  scale_y_continuous(labels = scales::comma_format()) +
  ggtitle("US Private Employment","Industry Portion of Total Employment") +
  labs(caption = "Source: Quarterly Census of Employment and Wages (QCEW)") +
  ylab("Annual Employment") +
  scale_fill_manual(labels = c(graph_names[1], graph_names[2], 
                               graph_names[3], 'Total \nEmployment'),
                    values = c(EFPcolors[3], EFPcolors[1], 
                               EFPcolors[2], EFPcolors[4])) +
  efp_theme
```


<!-- Text <http://rmarkdown.rstudio.com>.
**Bold** 
`example code`
-->
